Hydras \& Co.

### 2022-05-16 doc/ordinal-chapter.tex
revision book (<= p. 61)

Such a structure may be proved correct relatively to a bigger ordinal notation or to Schütte's or \gaia's model.

@@

\subsubsection{The \texttt{Comparison} and \texttt{Comparable} classes}

We use the operational class \texttt{Comparison} of comparison functions to define the \texttt{Comparable} class, contributed by Jérémy Damour and Théo Zimmermann, which allows us to apply generic lemmas and tactics about decidable strict orders.
The correctness of the comparison function is expressed through Stdlib's type 
\texttt{Datatypes.CompareSpec} as specialized by \texttt{Datatypes.CompSpec}.

@@

\begin{remark}
The infix notations \texttt{o<} and \texttt{o<=} are defined in order to make apparent the distinction between the various notation scopes that may co-exist in a same statement. So the infix \texttt{<} and \texttt{<=} are reserved to the natural numbers. In the mathematical formulas, however, we still use $<$ and $\leq$ for comparing ordinals.
\end{remark}

@@

\inputsnippets{ON_Omega/OmegaDefa, ON_Omega/OmegaDefb}

@@

\begin{itemize}
\item A \emph{least} element is a minorant (in the large sense) of the full set  on $A$,
\item $y$ is a \emph{successor} of $x$ if $x<y$ and there is no element between $x$ and $y$. We will also say that $x$ is a \emph{predecessor} of $y$.
\item $x$ is a \emph{limit} if $x$ is not a least element, and for any $y$ such that $yo<x$,
 there exists some $z$ such that $y<z<x$.
\end{itemize}

@@

\subsection{Arithmetic of \texorpdfstring{$\omega^2$}{omega^2}}

@@

\subsubsection{Multiplication}

@@

A case analysis on $i$ and $j$ allows us to
consider three cases :

@@

In order to build an instance of \texttt{ON}, we define a comparison function,  and prove its correctiness.

@@

Thus unicity of proofs of \texttt{Nat.ltb x0 (S n)}  comes from the decidability of
equality on type \texttt{bool}.
This is why we used the boolean function \texttt{Nat.ltb} instead of the inductive predicate \texttt{Nat.lt} in the definition of type \texttt{t $n$} (see page~\pageref{def: Finite-ord-type}).
For more information about this pattern, please look at the numerous mailing lists and 
FAQs on \coq{}).

@@

Finite ordinals are also formalized in \ssreflect/\mathcomp~\cite{MCB}. In Module~\href{../theories/html/gaia_hydras.ON_gfinite.html}{gaia\_hydras.ON\_gfinite}, we build an instance of class \texttt{ON}.

@@

The following class specifies a strict inclusion of segments. The notation \texttt{OA} describes a segment $[0,\alpha)$, and \texttt{OB} is a larger segment (which contains a notation for $\alpha$, whilst $\alpha$ is not represented in \texttt{OA}). We require also  that the comparison functions of the two notation systems are compatible.

\begin{figure}[h]
   \centering
   \begin{tikzpicture}[very thick, scale=0.6]
\begin{scope}[color=blue]
\node (A) at (0,0) {$A$};
\node(A0) at (2,0)[label=below:$0$]{$\bullet$};
\node(A1) at (3,0)[label=below:$1$]{$\bullet$};
\node(A2) at (4,0)[label=below:$2$]{$\bullet$};
\node (Adots) at (6,0) {$\ldots$};
\end{scope}
\begin{scope}[color=red]
\node (B) at (0,2) {$B$};
\node(B0) at (2,2)[label=above:$0$]{$\bullet$};
\node(B1) at (3,2)[label=above:$1$]{$\bullet$};
\node(B2) at (4,2)[label=above:$2$]{$\bullet$};
\node (Bdots) at (6,2) {$\ldots$};
\node (b) at (8,2) [label=above:$\omega$]{$\bullet$};
\node (bsucc) at (9,2) [label=above:$\omega+1$]{$\bullet$};
\node (Bdots2) at (10,2) {$\ldots$};
\end{scope}
\begin{scope}[color=red!50!blue]
\draw [->,thin] (A0) -- node [auto] {$\iota$} (B0);
\draw [->,thin] (A1) -- node [auto] {$\iota$} (B1);
\draw [->,thin] (A2) -- node [auto] {$\iota$} (B2);
\draw [->,thin] (Adots) -- node [auto] {$\iota$} (Bdots);
\end{scope}
\end{tikzpicture}
   \caption{\textcolor{blue}{$A$} is a sub-segment  of \textcolor{red}{$B$}}
   \label{fig:subsegment}
 \end{figure}

>>>>

Such a structure should be proved correct relatively to a bigger ordinal notation or to Schütte's or \gaia's model.

@@

\subsubsection{The \texttt{Comparable} class}

The \texttt{Comparable} class, contributed by Jérémy Damour and Théo Zimmermann, allows us to apply generic lemmas and tactics about decidable strict orders.
The correctness of the comparison function is expressed through Stdlib's type 
\texttt{Datatypes.CompareSpec} as specialized by \texttt{Datatypes.CompSpec}.

@@

\begin{remark}
The infix notations \texttt{o<} and \texttt{o<=} are defined in order to make apparent the distinction between the various notation scopes that may co-exist in a same statement. So the infix \texttt{<} and \texttt{<=} are reserved to the natural numbers. In mathematical formulas, however, we still use $<$ and $\leq$ for inequalities between ordinals.
\end{remark}

@@

\inputsnippets{ON_Omega/OmegaDefa}
\inputsnippets{ON_Omega/OmegaDefb}

@@

\begin{itemize}
\item A \emph{least} element is a minorant (in the large sense) of the full set  on $A$,
\item $y$ is a \emph{successor} of $x$ if $x<y$ and there is no element between $x$ and $y$. We will also say that $x$ is a \emph{predecessor} of $y$.
\item $x$ is a \emph{limit} if $x$ is not a least element, and for any $y$ such that $y<x$,
 there exists some $z$ such that $y<z<x$.
\end{itemize}

@@

\subsection{Arithmetic on \texorpdfstring{$\omega^2$}{omega^2}}

@@

\subsubsection{Finite multiplication}

@@

A case analysis on $i$ and $j$ leads  us to
consider three cases :

@@

In order to build an instance of \texttt{ON}, we define a comparison function, then prove its correctness.

@@

Thus unicity of proofs of \texttt{Nat.ltb x0 (S n)}  comes from the decidability of
equality on type \texttt{bool}.
This is why we used the boolean function \texttt{Nat.ltb} instead of the inductive predicate \texttt{Nat.lt} in the definition of type \texttt{t $n$} (see page~\pageref{def:Finite-ord-type}).
For more information about this pattern, please look at the numerous mailing lists and 
FAQs on \coq{}).

@@

\gaiasign Finite ordinals are also formalized in \ssreflect/\mathcomp~\cite{MCB}. In Module~\href{../theories/html/gaia_hydras.ON_gfinite.html}{gaia\_hydras.ON\_gfinite}, we build an instance of class \texttt{ON}.

@@

The following class specifies a strict inclusion of segments. The ordinal notation \texttt{OA} describes a segment $[0,\alpha)$, and \texttt{OB}  a larger segment (which contains a notation for $\alpha$, whilst $\alpha$ is not represented in \texttt{OA}) (like in Fig.~\ref{fig:subsegment}). We require also the comparison functions of the two notation systems to be compatible.

\begin{figure}[h]
   \centering
   \begin{tikzpicture}[very thick, scale=0.6]
\begin{scope}[color=blue]
\node (A) at (0,0) {$\omega$};
\node(A0) at (2,0)[label=below:$0$]{$\bullet$};
\node(A1) at (3,0)[label=below:$1$]{$\bullet$};
\node(A2) at (4,0)[label=below:$2$]{$\bullet$};
\node (Adots) at (6,0) {$\ldots$};
\end{scope}
\begin{scope}[color=red]
\node (B)  at (0,2) {$\omega+\omega$};
\node(B0) at (2,2)[label=above:$0$]{$\bullet$};
\node(B1) at (3,2)[label=above:$1$]{$\bullet$};
\node(B2) at (4,2)[label=above:$2$]{$\bullet$};
\node (Bdots) at (6,2) {$\ldots$};
\node (b) at (8,2) [label=above:$\omega$]{$\bullet$};
\node (bsucc) at (9,2) [label=above:$\omega+1$]{$\bullet$};
\node (Bdots2) at (10,2) {$\ldots$};
\end{scope}
\begin{scope}[color=red!50!blue]
\draw [->,thin] (A0) -- node [auto] {$\iota$} (B0);
\draw [->,thin] (A1) -- node [auto] {$\iota$} (B1);
\draw [->,thin] (A2) -- node [auto] {$\iota$} (B2);
\draw [->,thin] (Adots) -- node [auto] {$\iota$} (Bdots);
\end{scope}
\end{tikzpicture}
   \caption{\textcolor{blue}{$\omega$} is a sub-segment  of \textcolor{red}{$\omega+\omega$}}
   \label{fig:subsegment}
 \end{figure}
